---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
I have forked Roger's [GitHub repository](https://github.com/dnrichardson/RepData_PeerAssessment1) and therefore have the data for this assignment read to go. It's currently in a zip file, which we will now decompress.

```{r unzip}
## Check current working directory
getwd()
## List the files
list.files()
## Unzip the activity.zip file
unzip("activity.zip")
```

Now, let's read in the data. But first, let's check what the first 2 lines look like. 

```{r read lines}
## Read the first couple of lines
readLines(con = file("activity.csv", "r"), n = 2)
```

```{r read csv}
## File looks good and has a header. Read it in and specify the column classes
activity <- read.csv("activity.csv", header = TRUE, stringsAsFactors = FALSE, 
                     colClasses = c("integer", "Date", "integer"))
## Check the first lines
head(activity)

## Check the last lines
tail(activity)
```

Looks like there are a lot of NAs in the steps variable. Let's quickly check if we actually have step data

```{r}
sum(activity$steps, na.rm = TRUE)
```

Okay, looks good. Let's proceed. At this point I am going to load up dplyr and convert the activty 
dataframe to a tibble. 

```{r}
require(dplyr, quietly = TRUE, warn.conflicts = FALSE)

activity <- tbl_df(activity)
```


## What is mean total number of steps taken per day?
```{r mean steps per day}
## Try with tapply

tapply(activity$steps, activity$date, mean, na.rm = TRUE)

```

Ok. Mean steps per day are very low. This could mean we have a couch potato, or that the means are 
not a great reflection of daily activity. Let's do the median steps per day.

```{r median steps per day}

tapply(activity$steps, activity$date, median, na.rm = TRUE)

```

Wow. The median steps per day are ZERO! That doesn't seem very helpful. 

The means and medians output like this are a bit difficult to apprehend. Now, let's create a histogram of the total 
number of steps taken each day. We will use ggplot2. 

```{r hist of total steps per day}
## load ggplot2
require(ggplot2, quietly = TRUE)
require(scales, quietly = TRUE)

## Calculate total steps per day and convert to data frame
totSteps <- as.data.frame.table(tapply(activity$steps, activity$date, sum, na.rm = TRUE))

## Reconvert the date column to the class Date; note that the date variable was renamed to "Var1"
## by the as.data.frame.table function above
totSteps$Var1 <- as.Date(totSteps$Var1)

## Plot the histogram
ggplot(data = totSteps, aes(x = Var1)) + geom_histogram(aes(weights = Freq), binwidth = 1,
                                                        alpha = 0.5) + theme_bw() +
        labs(title = "Histogram of total steps per day") +
        labs(x = "Day", y = "Total number of Steps")

```

Alright, that looks fairly okay. I can live with this histogram and it took me suprisingly long to 
figure out how to plot it!


## What is the average daily activity pattern?
Make a time series plot (i.e. ðšðš¢ðš™ðšŽ = "ðš•") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

This will require me to calulate the average number of steps taken across all days for each interval. 

```{r means by intervals for all days}
## Use tapply again
meanStepsbyIntervals <- as.data.frame.table(tapply(activity$steps, activity$interval, mean, na.rm = TRUE)) %>%
        dplyr::rename(interval = Var1, mean = Freq)

## Check first lines
head(meanStepsbyIntervals)
```

Double check that I indeed calculated the mean for each interval. Check intervals 0 and 5 with dplyr:

```{r}
mean(filter(activity, interval == 0)$steps, na.rm = TRUE)
mean(filter(activity, interval == 5)$steps, na.rm = TRUE)
```

Yep, this all checks out. Let's move onto the time series plot, again using ggplot2.

```{r}

## Interval on the x axis and average steps taken on the y

ggplot(data = meanStepsbyIntervals, aes(x = interval, y = mean, group = 1)) + geom_line(na.rm = TRUE) +
        theme(axis.text.x = element_text(angle = 90, hjust = 1), text = element_text(size = 10))

```

Hmm.. ggplot2 is not producing what I want. The x-axis labels are **not cool**. I am going to try with plotly.

```{r fig.width = 8}
require(plotly, quietly = TRUE, warn.conflicts = FALSE)

xaxis <- list(title = "Interval",
              tickangle = 45
              )

plot_ly(meanStepsbyIntervals, x = ~interval, y = ~mean, type = "scatter", mode = "lines") %>%
        layout(title = "Mean steps per day across all days according to Interval", xaxis = xaxis)
```


Ahh, that is SOOO much better (and interactive, too)!

Now, onto the question of: 
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

From the graph above, the answer to this question is **interval 835, which has roughly 206 average steps**. 

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
