---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
I have forked Roger's [GitHub repository](https://github.com/dnrichardson/RepData_PeerAssessment1) and therefore have the data for this assignment read to go. It's currently in a zip file, which we will now decompress.

```{r unzip}
## Check current working directory
getwd()
## List the files
list.files()
## Unzip the activity.zip file if necessary
if (!file.exists("activity.csv")){
        unzip("activity.zip")
}
```

Now, let's read in the data. But first, let's check what the first 2 lines look like. 

```{r read lines}
## Read the first couple of lines
readLines(con = file("activity.csv", "r"), n = 2)
```

```{r read csv}
## File looks good and has a header. Read it in and specify the column classes
activity <- read.csv("activity.csv", header = TRUE, stringsAsFactors = FALSE, 
                     colClasses = c("integer", "Date", "integer"))
## Check the first lines
head(activity)

## Check the last lines
tail(activity)
```

Looks like there are a lot of NAs in the steps variable. Let's quickly check if we actually have step data

```{r}
sum(activity$steps, na.rm = TRUE)
```

Okay, looks good. Let's proceed. At this point I am going to load up dplyr and convert the activty 
dataframe to a tibble. 

```{r}
require(dplyr, quietly = TRUE, warn.conflicts = FALSE)

activity <- tbl_df(activity)
```


## What is mean total number of steps taken per day?

#### Calculate the mean number of steps per day using tapply.
```{r mean steps per day}
## Try with tapply

tapply(activity$steps, activity$date, mean, na.rm = TRUE)

```

Ok. Mean steps per day are very low. This could mean we have a couch potato, or that the means are 
not a great reflection of daily activity. Let's do the median steps per day.

#### Calculate the median of total number of steps taken per day.

```{r median steps per day}

tapply(activity$steps, activity$date, median, na.rm = TRUE)

```

Wow. The median steps per day are ZERO! That doesn't seem very helpful. 

The means and medians output like this are a bit difficult to apprehend. Now, let's create a histogram of the total number of steps taken each day. We will use ggplot2. 

```{r hist of total steps per day}
## load ggplot2
require(ggplot2, quietly = TRUE)
require(scales, quietly = TRUE)

## Calculate total steps per day and convert to data frame
totSteps <- as.data.frame.table(tapply(activity$steps, activity$date, sum, na.rm = TRUE))

## Reconvert the date column to the class Date; note that the date variable was renamed to "Var1"
## by the as.data.frame.table function above
totSteps$Var1 <- as.Date(totSteps$Var1)

## Plot the histogram
ggplot(data = totSteps, aes(x = Var1)) + geom_histogram(aes(weights = Freq), binwidth = 1,
                                                        alpha = 0.5) + theme_bw() +
        labs(title = "Histogram of total steps per day") +
        labs(x = "Day", y = "Total number of Steps")

```

Alright, that looks fairly okay. I can live with this histogram and it took me suprisingly long to 
figure out how to plot it!

## What is the average daily activity pattern?
Make a time series plot (i.e. 𝚝𝚢𝚙𝚎 = "𝚕") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

This will require me to calulate the average number of steps taken across all days for each interval. 

```{r means by intervals for all days}
## Use tapply again
meanStepsbyIntervals <- as.data.frame.table(tapply(activity$steps, activity$interval, mean, na.rm = TRUE)) %>%
        dplyr::rename(myinterval = Var1, mean = Freq)

## Check first lines
head(meanStepsbyIntervals)
```

Double check that I indeed calculated the mean for each interval. Check intervals 0 and 5 with dplyr:

```{r}
mean(filter(activity, interval == 0)$steps, na.rm = TRUE)
mean(filter(activity, interval == 5)$steps, na.rm = TRUE)
```

Yep, this all checks out. Let's move onto the time series plot, again using ggplot2.

#### Time series plot with ggplot2
```{r}

## Interval on the x axis and average steps taken on the y

ggplot(data = meanStepsbyIntervals, aes(x = as.numeric(as.character(myinterval)), y = mean, group = 1)) + geom_line(na.rm = TRUE) + labs(title = "Mean number of steps per day by interval", x = "interval",
                                                                                                                                         y = "mean steps per day")

```

Hmm.. The plot is nice, but it's hard to decipher what the maximum point is just from the plot. 
Let's try with an interactive graph by plotly. 

#### Interactive time series plot made with Plotly.
```{r fig.width = 8}
require(plotly, quietly = TRUE, warn.conflicts = FALSE)

xaxis <- list(title = "Interval",
              tickangle = 45
              )

plot_ly(meanStepsbyIntervals, x = ~myinterval, y = ~mean, type = "scatter", mode = "lines") %>%
        layout(title = "Mean steps per day across all days according to Interval", xaxis = xaxis)

```


Ahh, that is SOOO much better (and interactive, too)! Note: If you are not currently looking at the HTML
version of this file, you will not see the above plot! You must be looking at the knitted HTML document
to see the interactive plot above.

#### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

From the graph above, the answer to this question is **interval 835, which has roughly 206 average steps**. 

## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with 𝙽𝙰s)

```{r}
## Count up the total number of rows with NAs using base R
nrow(activity[activity$steps == "NA",])
```

```{r}
## Double check using dplyr
nrow(filter(activity, is.na(steps)))
```

Ok, so it looks like there are **2304 rows** with missing values. 

Let's now devise a simple strategy to impute the missing data. I'll use the mean of the 5 minute 
interval across all days to fill in the missing value at the interval in question. Hmm. I may 
need to write a function to do this. But first, I'll try the quick and dirty way.

```{r cbind}
## Simply cbind the the meanStepsbyIntervals to the activity dataframe
activity2 <- cbind(activity, meanStepsbyIntervals)

head(activity2)
```

Now, let's impute the missing values by simply checking if steps is NA and if it is, replacing that 
NA with the mean. 

```{r impute}
## Replace the NAs with the mean values
activity2[is.na(activity2$steps),]$steps <- activity2[is.na(activity2$steps),]$mean
head(activity2)
```

Now, I need to do the following:

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

For this, I can reuse the code I already created in the beginning of the document.

#### Calculate the mean steps per day

```{r mean steps per day 2}
## Try with tapply
tapply(activity2$steps, activity2$date, mean, na.rm = TRUE)
```

Looks like that worked. R is amazing! Note that by virtue of how I imputed the missing data, all
instances of missing data now have the exact same means (see 2012-10-01 and 2012-11-30 for example). Recall that I imputed the missing data by substituting in the mean of the interval across all days for the row with the NAs. It is more clear with the median values below.

```{r median steps per day 2}

tapply(activity2$steps, activity2$date, median, na.rm = TRUE)

```

Wow. The median steps per day are still ZERO! Except, it's important to note that by virtue of how
I imputed the missing data, all instances of missing data now have the exact same medians. 

#### Create histogram of total steps per day using ggplot

```{r hist of total steps per day 2}

## Calculate total steps per day and convert to data frame
totSteps <- as.data.frame.table(tapply(activity2$steps, activity2$date, sum, na.rm = TRUE))

## Reconvert the date column to the class Date; note that the date variable was renamed to "Var1"
## by the as.data.frame.table function above
totSteps$Var1 <- as.Date(totSteps$Var1)

## Plot the histogram
ggplot(data = totSteps, aes(x = Var1)) + geom_histogram(aes(weights = Freq), binwidth = 1,
                                                        alpha = 0.5) + theme_bw() +
        labs(title = "Histogram of total steps per day with imputed missing data") +
        labs(x = "Day", y = "Total number of Steps")

```

The impact of imputing missing data on the histogram is that the histogram has become more filled out
as the only gaps in the chart arise from days where the mean values for total steps taken is 0.

## Are there differences in activity patterns between weekdays and weekends?

Create a temporary variable, called "day", which stores the output of the weekdays() function. Then
create a new factor variable called, "weekday", that will have two levels, "weekday" and "weekend".
Note, I will be using the imputed data frame as there was no specification in the assingment instructions of which to use.

```{r weekdays}
## Create day variable
activity2$day <- weekdays(activity2$date)
head(activity2)

## Create weekday factor variable

activity2 <- activity2 %>% mutate(weekday = ifelse(day == "Saturday" | day == "Sunday", "weekend", ifelse(day != "Saturday" | day != "Sunday", "weekday")))

activity2$weekday <- factor(activity2$weekday, levels = c("weekday", "weekend"))

head(activity2)
```

Finally, we can make our last plot. This assignment was estimated to take 2 hours, but I've spent more like
5 hours on it.

```{r facet plot of average number of steps taken averaged across all weekday days or weekends }

## Create two new dataframes based on factor

weekdays <- activity2[activity2$weekday == "weekday",]
weekends <- activity2[activity2$weekday == "weekend",]

## Create two new dataframes of interval means
meanStepsbyIntervalsWeekdays <- as.data.frame.table(tapply(weekdays$steps, weekdays$interval, mean, na.rm = TRUE)) %>%
        dplyr::rename(myinterval = Var1, mean = Freq)

meanStepsbyIntervalsWeekends <- as.data.frame.table(tapply(weekends$steps, weekends$interval, mean, na.rm = TRUE)) %>%
        dplyr::rename(myinterval = Var1, mean = Freq)

## Add the weekend variable to facet by
meanStepsbyIntervalsWeekdays$weekday <- "weekday"
meanStepsbyIntervalsWeekends$weekday <- "weekend"

## rbind them
myIntervalData <- rbind(meanStepsbyIntervalsWeekdays, meanStepsbyIntervalsWeekends)

```

#### Facet plot of mean steps per day by interval 

```{r fig.width = 8}
## Interval on the x axis and average steps taken on the y

ggplot(data = myIntervalData, aes(x = as.numeric(as.character(myIntervalData$myinterval)), y = mean, group = 1)) + geom_line(na.rm = TRUE) +
        facet_grid(weekday ~.) + labs(x = "Interval", y = "Mean Steps per Day", title = "Mean Steps per Day by Interval and Weekday")

```

Indeed, it looks like activity levels are slightly higher in the intervals after 1000, but are lower in the intervals
below 1000. 


